Supplementary Information

### 2021-01-06 doc/manuscript/MEE_manuscript.Rmd
update climate and DBH results
#105

```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Dec.21)
#sameV_sameW: LDWp, NMt, NIOt, NIOp, SCBIp, SCt, ZOFp
#sameV_diffW: BCNMp, BCNMt, CBp, CBt, HFt, HKKp, LDWt, NMp, SCBIt, SCp, ZOFt
#diffV_diffW: HFp, HKKt
```
The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$)  exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs. `r all_climwin`).
In 7 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
In 11 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs. `r clim_sens`, `r GLS_HKK`, `r GLS_HF`), *climwin* identified different climate variables and different time windows (e.g., Fig. `r climwin_diffV_diffW`). 
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed).

```{r , eval=FALSE, echo=FALSE}
## Tallies for the paragraph below (as of Dec. 22)
#BCNM: 1 species with both, 2 with just P
#HKK: 3 species with both, 1 with just T
#SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
#LDW:  4 species with both,  2  with just P
#HF:  4 species with both
#ZOF:  3 species with both
#NIO:  1 species with both
#LT:  2 species with both
#CB:  7 species with both,  1 with just P (POTR)
#SC:  1 species with both
```
Within the context of the GLS model, which tested for first- and negative second- order linear effects of both a precipitation and a temperature variable, both a precipitation and a temperature variable were included in the top model for 80% of site-species combinations (n=37 of 46; Fig. `r clim_sens`). 
There were eight site-species combinations for which only a precipitation term was significant (2 at BCNM, 3 at SCBI, 2 at LDW, and 1 at CB), one for which only a temperature term was significant (CHTA at HKK), and none with no significant climate responses. 
Precipitation and temperature variables were rarely influential over the same time window (exception: LDW).
Below, we summarize the precipitation and temperature variables included in these models and their direction of response.

@@

```{r , eval=FALSE, echo=FALSE}
## Tallies for the  paragraph below (Dec.21)

@@

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo) 
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
#LDW: precip c June (1 mo), PET current June (1mo)
#HF: PDF pMay (1mo), PET cMar (1mo)
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
#NIO: wet cMay (1mo), TMP pJune (1mo)
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
#CB: P cApr-cJul(4mo), TMXpJul-cJul(13mo)
#SC: P pJun-pAug(3mo), TMXcJul (1mo)
```
Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
$PPT$ and $PDF$ were each identified as the top variable in four of the eight sites with both variables available, while $PPT$ was the only option at two (LT and CB).
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCNM and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
At three extratropical sites (LDW, NIO, and CB), the most influential windows were limited to the current spring and early summer, whereas optimal windows at two sites extended back to the previous fall (LT) or summer (SCBI).
Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but two species (*Tsuga canadensis* at HF and *Betula papyrifera* at NIO; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (*# of #* with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig. `r clim_sens`). 
At the riparian site (NIO), for which streamflow was included as a candidate variable in the precipitation group, May $PDF$ was selected as the strongest driver but had a modest negative effect on growth, indicating that water availability was not a significant limitation for *Betula papyrifera* at this site.

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses (*#*% of species; Fig. `r clim_sens`).
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites. 
$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCNM, respectively), noting that the effects of $T_{min}$ at BCNM were only marginally significant (Fig. `r clim_sens`).
Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), the previous growing season (NIO), or a 13-month window extending from past to current growing season (CB).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga Canadensis* at HF, the latter contrasting with a negative response of the other three species at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. **CREATE AN SI FIGURE!**).

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Dec.23)
n_spst_DBH_analyzed= 42 # n species with DBH analyzed: 42 (45 - 3)
n_spst_DBH_sig=34 #  species with sig first-order linear DBH term
n_spst_DBH_2nd_order=25
n_spst_DBH_2nd_order_neg=22
n_spst_BAI_DBH_best_model=41 #exception: ACRU at HF and 
n_spst_BAI_DBH_best_model_neg=40 #exception: PIPO at LT (reconstructed DBH down to only 15cm), which is neg but not significant
n_spst_BAI_DBH_best_model_2nd_neg=41
n_spst_dAGB_DBH_best_model=42 
```

@@

**Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).**
Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_neg` (exception: non-significant negative coefficient for *Pinus ponderosa* at LT, whose reconstructed DBH extended down to only 15cm within the time frame analyzed) and universally negative coefficients for $DBH^2$.
For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for $DBH$ and a negative coefficient for $DBH^2$.

>>>>

```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Jan 6)
#sameV_sameW: LDWp, NMt, NIOt, NIOp, SCBIp, SCt, ZOFp, CBt, CBp
#sameV_diffW: BCNMp, BCNMt, HFt, HKKp, LDWt, NMp, SCBIt, SCp, ZOFt
#diffV_diffW: HFp, HKKt
```
The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$)  exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs. `r all_climwin`).
In 9 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
In 9 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs. `r clim_sens`, `r GLS_HKK`, `r GLS_HF`), *climwin* identified different climate variables and different time windows (e.g., Fig. `r climwin_diffV_diffW`). 
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed).

```{r , eval=FALSE, echo=FALSE}
## Tallies for the paragraph below (as of Jan 6)
#BCNM: 1 species with both, 2 with just P
#HKK: 3 species with both, 1 with just T
#SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
#LDW:  4 species with both,  2  with just P
#HF:  4 species with both
#ZOF:  3 species with both
#NIO:  1 species with both
#LT:  2 species with both
#CB:  7 species with both,  1 with just P (PILO)
#SC:  1 species with both
```
Within the context of the GLS model, which tested for first- and negative second- order linear effects of both a precipitation and a temperature variable, both a precipitation and a temperature variable were included in the top model for 80% of site-species combinations (n=37 of 46; Fig. `r clim_sens`). 
There were eight site-species combinations for which only a precipitation term was significant (2 at BCNM, 3 at SCBI, 2 at LDW, and 1 at CB), one for which only a temperature term was significant (CHTA at HKK), and none with no significant climate responses. 
Precipitation and temperature variables were rarely influential over the same time window (exception: LDW).
Below, we summarize the precipitation and temperature variables included in these models and their direction of response.

@@

```{r , eval=FALSE, echo=FALSE}
## Tallies for the  paragraph below (Jan6)

@@

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo) 
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
#LDW: precip c June (1 mo), PET current June (1mo)
#HF: PDF pMay (1mo), PET cMar (1mo)
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
#NIO: wet cMay (1mo), TMP pJune (1mo)
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo)
#SC: P pJun-pAug(3mo), TMXcJul (1mo)
```
Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
$PPT$ and $PDF$ were each identified as the top variable in four of the eight sites with both variables available, while $PPT$ was the only option at two (LT and CB).
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCNM and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
At two extratropical sites (LDW, NIO), the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but two species (*Tsuga canadensis* at HF and *Betula papyrifera* at NIO; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (*# of #* with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig. `r clim_sens`). 
At the riparian site (NIO), for which streamflow was included as a candidate variable in the precipitation group, May $PDF$ was selected as the strongest driver but had a modest negative effect on growth, indicating that water availability was not a significant limitation for *Betula papyrifera* at this site.

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses (*#*% of species; Fig. `r clim_sens`).
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites. 
$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCNM, respectively), noting that the effects of $T_{min}$ at BCNM were only marginally significant (Fig. `r clim_sens`).
Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (*i.e.*, negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga Canadensis* at HF, the latter contrasting with a negative response of the other three species at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. **CREATE AN SI FIGURE!**).

@@

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Jan6)
n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
n_spst_DBH_sig=35 #  species with sig first-order linear DBH term
n_spst_DBH_2nd_order=27
n_spst_DBH_2nd_order_neg=22
n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF and 
n_spst_BAI_DBH_best_model_neg=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
n_spst_BAI_DBH_best_model_2nd=41
n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
```

@@

**Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).**
Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_neg` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed $DBH$s did not extend down to 0cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for $DBH$ and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
